Multitasking Interfaces on Touchscreen Mobile Devices

### 2015-12-09 sigchi-latex-proceedings 2/final-paper.tex
Edit Final Paper

%\pagenumbering{arabic}

@@

% End of preamble. Here it comes the document.
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%
\begin{document}

@@

\section{keywords}
Multitasking; Task switching interfaces; Mobile

@@

The mechanism by which users switch active apps varies between devices. Lists of applications, search methods, or gestures are common interactions to switch apps on smartphones. The most common way to switch apps is to open the application switching interface, or rather, the multitasking interface which varies by device. In the two major operating systems for smartphones, Android and iOS, each application is displayed in a list where the user can scroll through to select the desired application.

In this study, we aim to measure the app switching speed, accuracy, and perceived success of multiple methods for a user to switch to an already open app, regain cognitive state, perform a task, and then switch to another app in the Android operating system.

@@

In the study that is most closely related to this experiment, Warr and Chi [7] explored whether a "cards"-based mobile webpage switcher, which stacks open webpages like cards in the display, would result in faster webpage switching and less errors than a "pages"-based switcher, where the webpages are displayed as pages laid out next to each other. The research revealed that it is faster to switch webpages in the cards-based interface compared to the pages-based interface, but that there was no significant difference in error rates between the two interfaces [7]. Our research will be investigating both a cards-based application switcher and a pages-based switcher for speed and error rates.

@@

Given the cost to task completion time from switching apps, there have been many studies done on how to increase task resumption and completion time when multitasking. Leiva explored different techniques to improve task resumption time in a multi-tab browser environment [2], and Lottridge et al. investigated how the design of applications encourages or discourages multitasking behavior in general [4]. Leiva investigated the effects of a tool designed to remind users of their position on the screen prior to app switching as well as the last interacted element and cursor position, while Lottridge et al. looked at making users more aware of the time spent in "work" versus "non-work" categories of websites. Lottridge et al. found that by making users more aware of the passage of time, users had fewer webpage tabs open, fewer tab switches and shorter sessions within webpages [4], indicating that users were more efficient with their time spent on the web. Leiva found that the use of his tool improved task resumption time and task completion time.

Card and Henderson [1] observed that users interacting with an interface often organize tasks along a number of different dimensions, the most important of them being "locality" of tasks. The study found that users often form clusters of windows (apps) corresponding to a specific task. This insight led to development of interfaces that produce "virtual views" that allow users to focus their interactions within semantically meaningful clusters of windows. Modern smartphone multitasking interfaces sort opened apps in temporal order, but if a user is switching between two or three apps a time to perform a task, those apps in smartphone multitasking interfaces will be sorted together at the front of the list. If the apps are "clustered" to match with the user's mental model, this could result in faster and less error-prone application switching. In fact, Oliver et al. investigated a tool that indicated which windows in a multitasking interface were semantically related [6]. The use of this tool resulted in significantly more efficient task completion time, but it was noted that there was a negative effect when windows in the multitasking interface were reordered if the user interrupted the main task to switch to a different task [6]. This is what currently occurs in smartphone multitasking interfaces, which is why we intend to investigate the speed and accuracy of modern smartphone multitasking interfaces.

\section{Research Question and Hypotheses}
The purpose of this experiment is to determine how, if at all, different mobile multitasking interfaces affect accuracy in switching between applications as well as the time that it takes to switch applications. We will be investigating two different interfaces: a "stacked" interface where the pages that represent different opened apps overlap with each other, and a "non-stacked" interface where the pages are separated. \\

@@

The reasoning behind this is that while you can have more applications visible at any given time in a stacked layout, each application has a smaller target area to select, which translates into less accuracy.

\section{Methodology}
\section{Apparatus}
Two LG Nexus 5 phones were used as the apparatus for the experiment, one loaded with the stacked multitasking interface and the other with the non stacked interface. The stacked interface uses Android OS version 5.1 (Lollipop) while the non-stacked uses Android OS 4.4.4 (Kitkat). Both Android OS versions used were official Google releases and did not include any carrier or manufacturer modifications.

@@

\begin{figure}
 \includegraphics[width=\linewidth]{img/layout_no_cap.png}
 \caption{Left: the non-stacked card multitasking interface of Android 4.4 KitKat. Right: the stacked card interface of Android 5.0 Lollipop}
\end{figure}

\section{Participants}
Sixteen participants (ages 19-26) were recruited and did not receive monetary compensation for their participation. They all were currently pursuing an undergraduate degree at the University of Toronto. All participants met the criteria of previous experience with the Android operating system and no visual or hand dexterity disabilities. Participants were randomly assigned to begin with the stacked interface (n=8) or non-stacked interface (n=8). The study was carried out in a meeting room at the Mobile Application Development Lab (MADLab) which is located at the University of Toronto.

\section{Experimental Design}
The experiment had a one-way within-subject design with the multitasking interface (stacked and non-stacked) acting as a two-level independent variable. Order effects were counterbalanced by switching the order of testing with two groups of participants. The first group tested the stacked interface first, while the second group tested the non-stacked interface first. Time to completion and accuracy served as the dependent variables. Participants performed the set of tasks twice for each stacked and non-stacked multitasking interfaces with two different experimenters.

Randomizing participants into groups allowed us to minimize the variance any participants may have had in terms of greater experience or comfort with one interface over the other. Since the experimental tasks are repetitive, we were susceptible to the learning-effect. However, counterbalancing based on the two levels of the independent variable allowed us to account for it.

We designed the tasks to cover the three most common interactions with the multitasking interface. Firstly, minimal, where the participant’s interaction with the multitasking interface is limited to one or two switches. Secondly, a longer but a real world representative task where the participant switches between three or more apps. Lastly, repetitive switching, a common scenario where the participant repeatedly switches between two apps to achieve a specified goal. Combined with a within-subject design where the participant performs these tasks multiple times on each level gave us a very comprehensive set of data points to test our hypothesis.

\section{Tasks and Procedures}
The experimenter explained the purpose of the experiment to the participant and answered any general questions about the experiment. Each participant completed the entry survey (Appendix D) and signed the consent form (Appendix C). The experimenter showed the participant the experimental setup and described the tasks. For the setup, the experimenter confirmed that all apps needed for testing are pre-installed on the phone and are currently open, and that both phones are plugged into a computer and are being monitored. The experimenter confirmed that the participant knew how to open the multitasking interface and use the copy and paste function on the phone.

Each participant performed 3 blocks of tasks twice, once on KitKat and then on Lollipop. There were 6 trials in total per participant. The order of trials was balanced such that each participant performed the same set of tasks in the same order, but starting on a different device. Half the participants were randomly chosen to start on KitKat first, while the other half started on Lollipop first. Each participant was handed a printed sheet outlining all the tasks (see Appendix A). The participant read each block of tasks and executed them in the order listed on the sheet. Each block was preceded by a setup session for the experimenter to setup the device  (see Appendix B). The participant completed the tasks at his/her own pace. At the commencement of the experiment (6 trials), the participant completed an exit survey (see Appendix E). The whole experiment took about 20 minutes to complete per participant.

\section{Measures}
The dependent variables in this study were time spent in the multitasking interface, which corresponds to time taken to switch between apps, and the participants’ accuracy when switching apps. Time to complete (TTC) is measured as when the participant opens the multitasking interface to when the participant opens the target app. Accuracy is defined as when a participant does not open the target app in the task and instead either opens a different app or goes to the home screen. By measuring TTC and Accuracy, this will show which interface allows for fastest application switching.

\section{Data Collection}
Each participant completed a short questionnaire before the experiment to determine age, gender, handedness, as well as any prior experience with the device. Another questionnaire was completed after the experiment to determine preference and perceived speed and accuracy. The Android phones were connected to a MacBook Pro via micro-USB cable. Task completion time was logged using the Android Debug Bridge logging system. Errors were monitored by the experimenter and made note of. The task logs were parsed using a Python script, and the data was outputted in ANOVA readable format.

@@

\begin{center}
 \begin{tabular}{|c c c|}
 \hline
 Participant Number & Kitkat Speed & Lollipop Speed \\ [0.5ex]
 \hline\hline
 1 & 2.52 & 2.19\\
 \hline
 2 & 2.46 & 2.59 \\
 \hline
 4 & 2.28 & 2.14 \\
 \hline
 5 & 2.80 & 3.27 \\
 \hline
 6 & 2.76 & 2.17 \\
 \hline
 7 & 1.96 & 1.89 \\
 \hline
 8 & 2.02 & 1.79 \\
 \hline
 9 & 2.60 & 2.67 \\
 \hline
 10 & 2.06 & 1.17 \\
 \hline
 11 & 2.95 & 1.87 \\
 \hline
 12 & 2.28 & 1.71 \\
 \hline
 13 & 2.88 & 1.85 \\
 \hline
 14 & 1.92 & 2.72 \\
 \hline
 15 & 2.05 & 2.46  \\
 \hline
 16 & 2.69 & 3.03 \\
 \hline
  17 & 1.73 & 2.00 \\
 \hline \hline
 Avg & 2.37 & 2.22 \\

 \hline
\end{tabular}
\end{center}

Given that this was a within-subjects test where we had 1 independent variable, a one-way with one within-subjects factor ANOVA test was carried out.
The mean task completion times was 2.37 seconds for KitKat (non-stacked).  This was 6.7\% slower than the mean of Lollipop (stacked) which was 2.22 seconds. The difference was not statistically significant

@@

\begin{center}
 \begin{tabular}{|c c c|}
 \hline
 Participant Number & Kitkat Errors & Lollipop Errors \\ [0.5ex]
 \hline\hline
 1 & 1 & 1\\
 \hline
 2 & 10 & 5\\
 \hline
 4 & 0 & 0 \\
 \hline
 5 & 1 & 3 \\
 \hline
 6 & 4 & 1 \\
 \hline
 7 & 0 & 2 \\
 \hline
 8 & 0 & 1 \\
 \hline
 9 & 10 & 4 \\
 \hline
 10 & 1 & 0 \\
 \hline
 11 & 2 & 2 \\
 \hline
 12 & 3 & 3 \\
 \hline
 13 & 1 & 1 \\
 \hline
 14 & 0 & 0 \\
 \hline
 15 & 1 & 0  \\
 \hline
 16 & 2 & 1 \\
 \hline
  17 & 2 & 1 \\
 \hline \hline
 Avg & 2.37 & 1.56 \\

 \hline
\end{tabular}
\end{center}

The mean error rate for was 2.37 total errors for Kitkat and 1.56 total errors per participant for Lollipop. However, as there was substantial variation in the observations across participants, the difference was not statistically significant as revealed by the analysis of variance. Once again, a one-way with one within-subjects factor ANOVA test was carried out
\begin{center}
$(F_{1, 15} = 2.19, ns )$
\end{center}

The exit survey questions relating to the participant’s perceived accuracy and speed were converted to an ordinal scale between 1 and 5. Participants believed they were slightly faster on Lollipop (3.92 average response versus 3.76). Participants perceived they were more accurate on Kitkat (4.38 average response versus 4.07).

\section{Discussion}
While the differences between speed and accuracy are statistically insignificant, a discussion can still take place. The stacked interface was perceived by the participants to be faster, which held when examining the data. This is inline with previous literature like the work of Warr and Chi [7]. However, this begs the question: why is stacked interface perceived as quicker? It’s possible this is due to the fact that the motion of the cards is toward the user rather than down. In addition, with the stacked interface the user can control the sensitivity of the scroll by scrolling from higher up on the screen. This allows users to more quickly traverse a long list of apps at the cost of some granularity of motion.

A stacked interface may only provide nominal speed and accuracy increases when compared to a non-stacked interface. As a result, future research should focus on alternate means to facilitate app switching as opposed to the conventional historical listing of applications. Throughout all mobile app switching interfaces, almost exclusively, the “most recently used” rule is used to sort applications in the list. Future research should look into using different sorting schemes to predict which application a user will need next. Perhaps machine learning could be applied to predict what application may be used next as is being tested in Apple’s new recents menu.

@@

Overall, the study shows that the differences between the two interfaces are statistically insignificant for accuracy and speed. The stacked card interface was rated as generally more pleasant and quicker, which could be a reason Android changed the interface.

\section{Conclusion}

>>>>

\pagenumbering{arabic}

@@

% For shared affiliation
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%

% End of preamble. Here it comes the document.
\begin{document}

@@

\section{keywords}
Multitasking; Task switching interfaces; Touchscreen; Mobile

@@

The mechanism by which users switch active apps varies between devices. Lists of applications, search methods, or gestures are just some of the interactions to switch apps on smartphones. The most common way to switch apps is to open the application switching interface, or rather, the multitasking interface which varies by device. In the two major operating systems for smartphones, Android and iOS, each application is displayed in a list where the user can scroll through to select the desired application.

In this study, we aim to measure the app switching speed, accuracy, and perceived success of multiple methods for a user to switch to an already open app, perform a task, and then switch to another app in the Android operating system.

@@

@@

Given the cost to task completion time from switching apps, there have been many studies done on how to increase task resumption and completion time when multitasking. Leiva [2] explored different techniques to improve task resumption time in a multi-tab browser environment, and Lottridge et al. [4] investigated how the design of applications encourages or discourages multitasking behavior in general. Leiva investigated the effects of a tool designed to remind users of their position on the screen prior to app switching as well as the last interacted element and cursor position, while Lottridge et al. looked at making users more aware of the time spent in "work" versus "non-work" categories of websites. Lottridge et al. found that by making users more aware of the passage of time, users had fewer webpage tabs open, fewer tab switches and shorter sessions within webpages [4], indicating that users were more efficient with their time spent on the web. Leiva found that the use of his tool improved task resumption time and task completion time.

Card and Henderson [1] observed that users interacting with an interface often organize tasks along a number of different dimensions, the most important of them being "locality" of tasks. The study found that users often form clusters of windows corresponding to a specific task. This insight led to development of interfaces that produce "virtual views" that allow users to focus their interactions within semantically meaningful clusters of windows. Modern smartphone multitasking interfaces sort opened apps in chronological order, but if a user is switching between two or three apps a time to perform a task, those apps in smartphone multitasking interfaces will be sorted together at the front of the list. If the apps are "clustered" to match with the user's mental model, this could result in faster and less error-prone application switching. In fact, Oliver et al. investigated a tool that indicated which windows in a multitasking interface were semantically related [6]. The use of this tool resulted in significantly more efficient task completion time, but it was noted that there was a negative effect when windows in the multitasking interface were reordered if the user interrupted the main task to switch to a different task [6]. This is what currently occurs in smartphone multitasking interfaces, which is why we intend to investigate the speed and accuracy of modern smartphone multitasking interfaces.

In the study that is most closely related to this experiment, Warr and Chi [7] explored whether a "cards"-based mobile webpage switcher, which stacks open webpages like cards in the display, would result in faster webpage switching and less errors than a "pages"-based switcher, where the webpages are displayed as pages laid out next to each other. The research revealed that it is faster to switch webpages in the cards-based interface compared to the pages-based interface, but that there was no significant difference in error rates between the two interfaces [7]. These authors were specifically investigating a web browser switching interface and that was across two different operating systems. Our experiment is on one operating system, and focuses on speed and error rate difference between two common multitasking interfaces.

\section{Research Question and Hypotheses}
The purpose of this experiment is to determine how, if at all, different mobile multitasking interfaces affect accuracy in switching between applications as well as the time that it takes to switch applications. We will be investigating two different interfaces: a stacked interface where the pages that represent different opened apps overlap with each other, and a non-stacked interface where the pages are separated. \\

@@

The reasoning behind this is that while there can be more applications visible at any given time in a stacked layout, each application has a smaller target area to select, which translates into less accuracy.

\section{Methodology}
\subsection{Apparatus}
Two LG Nexus 5 phones were used as the apparatus for the experiment, one loaded with the stacked multitasking interface and the other with the non stacked interface. The stacked interface uses Android OS version 5.1 (Lollipop) while the non-stacked uses Android OS 4.4.4 (KitKat). Both Android OS versions used were official Google releases and did not include any carrier or manufacturer modifications.

@@

\begin{figure}
 \includegraphics[width=\linewidth]{img/layout_no_cap.png}
 \caption{Left: The non-stacked multitasking interface of Android 4.4 KitKat. Right: The stacked interface of Android 5.0 Lollipop.}
\end{figure}

\subsection{Participants}
Sixteen participants (ages 19-26) were recruited and did not receive monetary compensation for their participation. They all were currently pursuing an undergraduate degree at the University of Toronto. All participants met the criteria of previous experience with the Android operating system and no visual or hand dexterity disabilities. Participants were randomly assigned to begin with the stacked interface (n=8) or non-stacked interface (n=8). The study was carried out in a meeting room at the Mobile Application Development Lab (MADLab) which is located at the University of Toronto.

\subsection{Experimental Design}
The experiment had a one-way within-subject design with the multitasking interface (stacked and non-stacked) acting as a two-level independent variable. Order effects were counterbalanced by switching the order of testing with two groups of participants. The first group tested the stacked interface first, while the second group tested the non-stacked interface first. Time to completion and accuracy served as the dependent variables. Participants performed the set of tasks twice for each stacked and non-stacked multitasking interfaces with two different experimenters.

Randomizing participants into groups allowed us to minimize the variance any participants may have had in terms of greater experience or comfort with one interface over the other. Since the experimental tasks are repetitive, the participants were susceptible to the learning-effect. However, counterbalancing based on the two levels of the independent variable allowed us to account for it.

We designed the tasks to cover the three most common interactions with the multitasking interface. Firstly, minimal, where the participant’s interaction with the multitasking interface is limited to one or two switches. Second, a longer but real-world representative task where the participant switched between three or more apps. Last, repetitive switching, a common scenario where the participant repeatedly switched between two apps to achieve a specified goal. Combined with a within-subject design where the participant performs these tasks multiple times on each level gave us a  comprehensive set of data points to test the hypotheses.

\subsection{Tasks and Procedures}
The experimenter explained the purpose of the experiment to the participant and answered any general questions about the experiment. Each participant completed the entry survey and signed the consent form. The participants were free to withdraw at any point during the experiment. The experimenter showed the participant the experimental setup and described the tasks. For the setup, the experimenter confirmed that the phone was set up for each task, and that both phones were plugged into a computer and were being monitored. The experimenter confirmed that the participant knew how to open the multitasking interface and use the copy and paste function on the phone.

Each participant performed three blocks of tasks twice, once on each multitasking interface. There were six trials in total per participant. Each participant was handed a sheet outlining all the tasks. The participant read each block of tasks and executed them in the order listed on the sheet. Each block was preceded by a short break for the experimenter to setup the device. The participant completed the tasks at his/her own pace. At the commencement of the experiment, the participant completed an exit survey. The whole experiment took about 20 minutes to complete per participant.

\subsection{Measures}
The dependent variables in this study were time spent in the multitasking interface, which corresponds to time taken to switch between apps, and the  accuracy when switching apps. Time to completion is measured as when the participant opens the multitasking interface to when the participant opens the target app. Accuracy is defined as when a participant does not open the target app in the task and instead either opens a different app or goes to the home screen. Measuring time to completion and accuracy will show which interface allows for fastest and least error-prone application switching.

\subsection{Data Collection}
Each participant completed a short questionnaire before the experiment to determine age, gender, handedness, prior experience with Android, whether they owned an Android phone, and which OS version they were most comfortable with. Another questionnaire was completed after the experiment to determine perceived speed and accuracy with the two multitasking interfaces, as well as general preference for one interface over another. During the experiment, the Android phones were connected to a MacBook Pro via micro-USB cable. Task completion time was logged using the Android Debug Bridge logging system. Errors were monitored by the experimenter and made note of. The task logs were parsed using a Python script, and the data was outputted in ANOVA readable format.

@@

%\begin{center}
% \begin{tabular}{|c c c|}
% \hline
% Participant Number & Kitkat Speed & Lollipop Speed \\ [0.5ex]
% \hline\hline
% 1 & 2.52 & 2.19\\
% \hline
% 2 & 2.46 & 2.59 \\
% \hline
% 4 & 2.28 & 2.14 \\
% \hline
% 5 & 2.80 & 3.27 \\
% \hline
% 6 & 2.76 & 2.17 \\
% \hline
% 7 & 1.96 & 1.89 \\
% \hline
% 8 & 2.02 & 1.79 \\
% \hline
% 9 & 2.60 & 2.67 \\
% \hline
% 10 & 2.06 & 1.17 \\
% \hline
% 11 & 2.95 & 1.87 \\
% \hline
% 12 & 2.28 & 1.71 \\
% \hline
% 13 & 2.88 & 1.85 \\
% \hline
% 14 & 1.92 & 2.72 \\
% \hline
% 15 & 2.05 & 2.46  \\
% \hline
% 16 & 2.69 & 3.03 \\
% \hline
%  17 & 1.73 & 2.00 \\
% \hline \hline
% Avg & 2.37 & 2.22 \\
%
% \hline
%\end{tabular}
%\end{center}

Given that this was a within-subjects test with one independent variable, a one-way with one within-subjects factor ANOVA test was carried out.
The mean task completion times was 2.37 seconds for the non-stacked interface (KitKat). This was 6.7\% slower than the mean of stacked interface (Lollipop) which was 2.22 seconds. The difference was not statistically significant.

@@

%\begin{center}
% \begin{tabular}{|c c c|}
% \hline
% Participant Number & Kitkat Errors & Lollipop Errors \\ [0.5ex]
% \hline\hline
% 1 & 1 & 1\\
% \hline
% 2 & 10 & 5\\
% \hline
% 4 & 0 & 0 \\
% \hline
% 5 & 1 & 3 \\
% \hline
% 6 & 4 & 1 \\
% \hline
% 7 & 0 & 2 \\
% \hline
% 8 & 0 & 1 \\
% \hline
% 9 & 10 & 4 \\
% \hline
% 10 & 1 & 0 \\
% \hline
% 11 & 2 & 2 \\
% \hline
% 12 & 3 & 3 \\
% \hline
% 13 & 1 & 1 \\
% \hline
% 14 & 0 & 0 \\
% \hline
% 15 & 1 & 0  \\
% \hline
% 16 & 2 & 1 \\
% \hline
%  17 & 2 & 1 \\
% \hline \hline
% Avg & 2.37 & 1.56 \\
%
% \hline
%\end{tabular}
%\end{center}

The mean error rate for was 2.37 errors for the non-stacked interface and 1.56 errors for the stacked interface. However, as there was substantial variation in the errors across participants, the difference was not statistically significant as revealed by the analysis of variance. Once again, a one-way with one within-subjects factor ANOVA test was carried out.
\begin{center}
$(F_{1, 15} = 2.19, ns )$
\end{center}

The exit survey questions relating to the participant’s perceived accuracy and speed were converted to an ordinal scale between 1 and 5. Participants believed they were slightly faster on the stacked interface (3.92 average response versus 3.76). Participants perceived they were more accurate on the non-stacked interface (4.38 average response versus 4.07).

\section{Discussion}
While the trends in the speed and accuracy measurements of our participants was statistically insignificant between the two interfaces, a discussion can still take place. The stacked interface was perceived by the participants to be faster, which held when examining the data. This is inline with previous research, such as the work of Warr and Chi [7]. However, a follow-up question is presented: why is stacked interface perceived as quicker? It?s possible this is due to the fact that the motion of the cards is toward the user rather than down. In addition, with the stacked interface the user can control the sensitivity of the scroll by scrolling from higher up on the screen. This allows users to more quickly traverse a long list of apps at the cost of some granularity of motion.

\section{Conclusion}
Overall, the experiment showed that the differences between the two interfaces are statistically insignificant for accuracy and speed. The stacked  interface was rated as generally more pleasant and quicker, which could be a reason Google updated Android?s default multitasking interface.

@@

A stacked interface may only provide nominal speed and accuracy increases when compared to a non-stacked interface. As a result, future research should focus on alternate means to facilitate app switching as opposed to the conventional chronological list of applications. Throughout all mobile multitasking interfaces, almost exclusively, the ?most recently used? rule is used to display applications in the list. Future research could look into using different sorting schemes to predict which application a user will need next. Perhaps machine learning could be applied to predict what application may be used next as is being tested in Apple?s new recents menu.
